Predicting drug-induced perturbation effects is challenging due to complex gene networks and limited mechanistic knowledge. Graph neural networks (GNNs) are well suited to this problem because their node-and-edge structure mirrors biological interaction graphs and can integrate multimodal inputs. In this thesis, we build a GNN model that combines (1) a gene interaction graph from the Omnipath dataset, which contains gene interaction and regulation data, (2) cell-specific baseline gene expression from CMAP LINCS L1000 Level 3 beta, and (3) drug information from target annotations and molecular fingerprints. The model predicts gene expression changes after drug treatment.

We evaluate the approach on public perturbation data using cold-drug and cold-cell splits, where test drugs or cell lines are unseen during training. We also run sanity checks and baseline comparisons, including an MLP model, to quantify the contribution of graph structure and drug inputs. Under single-dose, single-time settings, the GNN improves cold-cell performance by 24.4\% over the MLP baseline. Additional ablation tests remove or shuffle drug inputs to assess how much the model depends on drug targets and fingerprints, providing a clearer view of which signals drive performance. 
Furthermore, by analyzing the attention weights in the GAT layers, we gain insight into how drug information propagates through the gene network and which genes play key roles in mediating drug effects. This interpretability helps reveal the molecular mechanisms underlying the model's predictions and highlights biologically meaningful pathways influenced by drug perturbation.
Finally, we show that integrating interaction graphs with cell context and drug annotations improves generalization under cold-start conditions.
This work demonstrates that integrating biological interaction networks with graph neural networks substantially improves the prediction of drug perturbation effects, especially for unseen drugs and samples, and provides new insights beyond traditional machine learning approaches.